---
title: "First hands-on for the DMI course"
author: "Wael Badr and Xareni Reyes Soto"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Analysis of the Heart Disease Dataset 
Load the data from
[here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). 
The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Perform an EDA on the dataset

As we saw in class, the major tasks during EDA are data cleaning, data integration, data reduction and data transformation. We will start by cleaning the data: we will handle missing data, check if there are inconsistencies with the data types, and identify and remove outliers.
Data integration is not necessary for this exercise, since we are working with only one file.

TODO: are data reduction (dimensionality reduction, numerosity reduction and datacompression) and data transformation (normalizing, scaling) applicable to this exercise?
```{r}
library(outliers)
library(rmarkdown)
# Read the table
data_frame <- read.table("/home/xrs/projects-ubuntu/MBI_data_mining/assignment-1-xreyesso/data/heart_disease_dataset.csv", header=TRUE, sep=" ")

# The column patient_id works great as entity identifier. Rearrange the data so that this column is now the first
data_frame <- data_frame[, c(ncol(data_frame), 1:(ncol(data_frame) - 1))]

#Data Cleaning
# 1. Check the data types
# 2. Check for missing values
# 3. Check for duplicate rows
# 4. Check outliers and remove them
# 5. Statistics summary

# 1. Check the data types
str(data_frame)

# We notice that columns ca and thal have type 'char', so we need to convert them to integers for further processing and visualization
data_frame$ca <- as.numeric(data_frame$ca) 
data_frame$thal <- as.numeric(data_frame$thal)

# We verify that the columns have the correct data types after the conversion
str(data_frame)

# 2. We check for missing values
missing_counts <- colSums(is.na(data_frame))
print(missing_counts)

# There 4 'NA' values in ca and 2 'NA' values in thal. We choose to delete them. For this, we replace data_frame with a new version of the table with the NA rows removed
data_frame <-  na.omit(data_frame)

# 3. Check if there are any duplicate rows in the table
any(duplicated(data_frame))     # returns FALSE

# 4. We proceed to check if there are outliers using the IQR method: outliers as defined as values less than Q1 - 1.5*IQR or greater than Q3 + 1.5*IQR

# Compute the 1st and 3rd quartiles (Q1 and Q3), using the apply function by columns
Q1 <- apply(data_frame, 2, quantile, 0.25, na.rm=TRUE)
Q3 <- apply(data_frame, 2, quantile, 0.75, na.rm=TRUE)

# Compute the Interquartile range (IQR)
IQR <- Q3 - Q1

# Identify rows that contain at least one outlier, we use the apply function by rows to identify them
outlier_rows <- apply(data_frame, 1, function(x) any(x < (Q1 - 1.5*IQR) | x > (Q3 + 1.5*IQR)))

# Store outlier rows in a separate data frame
outliers_df <- data_frame[outlier_rows, ]

# Remove outliers by doing data_frame <- data_frame[! ... , ]
data_frame <- data_frame[!outlier_rows, ]

# 5. Summary statistics
summary(data_frame)

```
Now that we cleaned the data, we proceed to do the Exploratory Data Analysis. We used the steps provided in https://www.kaggle.com/code/georgyzubkov/heart-disease-exploratory-data-analysis as guide.
```{r}
#XRS 1. Check the dataset for gaps in the data
# Create a data frame for the missing values for plotting
df_missing <- data.frame(Variable = names(missing_counts), Missing_Count = missing_counts)
# Load ggplot2
library(ggplot2)

# Plot
ggplot(df_missing, aes(x = reorder(Variable,-Missing_Count), y = Missing_Count)) +
  geom_bar(stat = "identity", fill = "darkblue", alpha = 0.7)+
  coord_flip() + # We flip the graph for improved readability
  labs(title = "Check for gaps in our data",
       x = "Variables",
       y = "Missing values counts") +
  theme_minimal()
```
We do have some missing values, but we removed them in the cleaning steps above.
```{r}
#XRS 2. Look at the distribution of the target variable: heart disease
num_counts <- as.data.frame(table(data_frame$num))

# Plot
ggplot(num_counts, mapping = aes(x=Var1, y=Freq)) +
  geom_bar(stat = "Identity", fill = "#6BACD0") +
  labs(title = "Distribution of our target variable",
       x = "num",
       y = "Count") +
  theme_minimal()
```
The sample seems to be unbalanced, with the majority of the patients being in the category 0 .
```{r}
#XRS 3. Indicate what are the numerical variables and which categorical


#XRS 4. Look at unique values in the categorical objects


#WB 5. Analyze numerical variables (visualizing continuous columns) -> plot them -> does any of them seem to have a normal distribution?
#WB 6. Look at the distribution of number of people with heart disease from various factors
#WB 7. Look at the distribution of categorical variables depending on gender
#WB 8. Look at outliers in numerical values depending on several factors 
#CONCLUSION ---> e.g. from all the graphs presented, it can be concluded that e.g. "alcohol consumption and smoking are not the main factors in heart disease"


```


## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r}

library(corrplot)
library(dplyr)
# Visualize Correlation Matrix
# Select relevant columns
selected_data <- data_frame[, c("num", "oldpeak", "age", "trestbps", "chol", "thalach")]

# Ensure columns are numeric
selected_data <- selected_data %>%
  mutate(across(everything(), as.numeric))

#correlation matrix with the numeric measurements
cor_matrix <- cor(selected_data, method = "spearman")
corrplot(cor_matrix, method = "color", type = "full", addCoef.col = "black",
         tl.col = "black", tl.srt = 45, number.cex = 0.8)

```


# 2. Difference in mortality rates in hospitalized COVID-19 patients 
Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}
library(readxl)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)

table <- read_excel("/home/xrs/Documents/02_DMI/Hands-on_1/Table 1.XLSX")

# Figure 1, part 1
# Standardize column names
table <- table %>%
  rename_with(~ gsub("\\s+", "_", .)) %>%
  rename_with(tolower)

# Step 1: Remove exact duplicate rows
table_unique <- table %>% distinct()

table_unique <- table_unique %>%
  mutate(age_group = cut(age, breaks = seq(20, 100, by = 10), right = FALSE, include.lowest = TRUE))

# Step 4: Create the histogram with adjustments for styling
ggplot(data = table_unique, mapping = aes(x = age)) +
  geom_histogram(
    binwidth = 10,
    color = "black",
    fill = "#E0EEEE",
    boundary = 20
  ) +
  scale_x_continuous(breaks = seq(20, 100, by = 10), limits = c(20, 100)) +
  labs(
    x = "Age (years)",
    y = "Frequency (n)",
    title = "Age"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

# Figure 1, part 2
# Define the table data
data <- matrix(c("G1", "-",  "-",   "-",
                 "G2", "-/+", "+/-", "-",
                 "G3", "+",   "-",   "+",
                 "G4", "-/+", "+",   "+"),
               nrow = 4, ncol = 4, byrow = TRUE)

colnames <- c("Clinical\nclassification", "NIV", "AMV", "ARDS")

# Create the tableGrob
table <- tableGrob(data,
                   rows = NULL, # No row names
                   cols = colnames)

# Define the theme with alternating row colors and centered text
theme <- ttheme_minimal(
  core = list(
    fg_params = list(hjust = 0.5, x = 0.5, fontsize = 12), # Center-align and font size
    bg_params = list(fill = c(rep(c("white", "lightgray"), length.out = nrow(data))), 
                     col = "white") # Alternating row colors with white gridlines
  ),
  colhead = list(
    fg_params = list(fontsize = 14, fontface = "bold"), # Header style
    bg_params = list(fill = "lightgray", col = "white") # Header background
  )
)

# Recreate the table with the theme
table <- tableGrob(data, rows = NULL, cols = colnames, theme = theme)

# Add white lines (gridlines) between rows and columns
table$grobs <- lapply(table$grobs, function(grob) {
  if (inherits(grob, "rect")) {
    grob$gp$col <- "white" # Set gridline color to white
    grob$gp$lwd <- 2       # Set line width
  }
  return(grob)
})

# Add a title
title <- textGrob("Definition of the clinical classification",
                  gp = gpar(fontsize = 14, fontface = "bold"))

# Combine title and table into a single plot with closer spacing
grid.newpage()
grid.arrange(table, ncol = 1) # I wasnt able to add the title without messing up the figure

```


## Reproduce Figure 2 from the publication
but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(pheatmap)

# 1. Read the data from the Excel document
raw_data <- read_excel("/home/xrs/Documents/02_DMI/Hands-on_1/Table 2.XLSX", 1)

# View the first few rows
head(raw_data)

# The first row contains a header
colnames(raw_data) <- raw_data[1, ] # Set the first row as column names
clean_data <- tail(raw_data,-1)      # Remove the first row

# For more clarity, make sure the first column is named "Patient ID"
colnames(clean_data)[1] <- "Patient_ID"

# Check data types of the data frame
str(clean_data)           # All columns have type 'char'

# 2. Transform cytokine data to numeric type for further processing
numeric_columns <- colnames(clean_data)[2:ncol(clean_data)]                    # Select columns except for the first one (Patient ID)
clean_data[numeric_columns] <- lapply(clean_data[numeric_columns], as.numeric) # Convert to numeric type, use the lapply function to apply as.numeric to each element of the numeric_columns
# Verify the columns have the correct type
str(clean_data)

# 3. Handle multiple measurements per patient
# First of all, remove leading/trailing spaces in Patient_ID
clean_data$Patient_ID <- trimws(clean_data$Patient_ID)

# Then, fill in the missing Patient_ID values for rows with 'NA' using the fill() function, for the 
# data aggregation to work correctly
clean_data <- clean_data %>%
  fill(Patient_ID, .direction = "down")

# Some patients have multiple measurements, taken on different days. To aggregate their cytokine levels, we choose to replace multiple measurements by their mean
aggregated_data <- clean_data %>%
  group_by(Patient_ID) %>%
  summarise(across(everything(), ~mean(.x, na.rm = TRUE)))

# Scale each cytokine between 0 (min) and 1 (max)
min_max_norm <- function(x) {
  return((x - min(x, na.rm = TRUE))/(max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))
}

# Apply normalization to all cytokine columns (except for Patient_ID)
normalized_data <- aggregated_data %>%
  mutate(across(-Patient_ID, min_max_norm))

# Only keep the first 14 columns of the data frame, except column Day (which contain the cytokine measurements we are interested in plotting)
new_normalized_data <- select(normalized_data, c(1,3:14))

# Create heatmap
# Convert to numeric matrix for heatmap
data_matrix <- as.matrix(new_normalized_data[,-1]) # Exclude patient ID

pheatmap(t(data_matrix),
         color = colorRampPalette(c("#FDF5EB", "#FD9C50","#8D2A08"))(100),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         fontsize_row = 10,
         main = "Cytokine Expression Heatmap (Relative Expression)")

```


```{r}

```


```{r}
```

## Improve figure 2 of the publication
Add a second annotation with information of deathm and a third one with information of gender

```{r}

```


# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
