---
title: "First hands-on for the DMI course"
author: "Wael Badr and Xareni Reyes Soto"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Analysis of the Heart Disease Dataset 
Load the data from
[here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). 
The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Perform an EDA on the dataset

As we saw in class, the major tasks during EDA are data cleaning, data integration, data reduction and data transformation. We will start by cleaning the data: we will handle missing data, check if there are inconsistencies with the data types, and identify and remove outliers.
Data integration is not necessary for this exercise, since we are working with only one file.

TODO: are data reduction (dimensionality reduction, numerosity reduction and datacompression) and data transformation (normalizing, scaling) applicable to this exercise?
```{r}
library(outliers)
library(rmarkdown)

# Read the table
data_frame <- read.table("data/heart_disease_dataset.csv", header=TRUE, sep=" ")

# The column patient_id works great as entity identifier. Rearrange the data so that this column is now the first
data_frame <- data_frame[, c(ncol(data_frame), 1:(ncol(data_frame) - 1))]

#Data Cleaning
# 1. Check the data types
# 2. Check for missing values
# 3. Check for duplicate rows
# 4. Check outliers and remove them
# 5. Statistics summary

# 1. Check the data types
str(data_frame)

# We notice that columns ca and thal have type 'char', so we need to convert them to integers for further processing and visualization
data_frame$ca <- as.numeric(data_frame$ca) 
data_frame$thal <- as.numeric(data_frame$thal)

# We verify that the columns have the correct data types after the conversion
str(data_frame)

# 2. We check for missing values
missing_counts <- colSums(is.na(data_frame))
print(missing_counts)

# There 4 'NA' values in ca and 2 'NA' values in thal. We choose to delete them. For this, we replace data_frame with a new version of the table with the NA rows removed
data_frame <-  na.omit(data_frame)

# 3. Check if there are any duplicate rows in the table
any(duplicated(data_frame))     # returns FALSE

# 4. We proceed to check if there are outliers using the IQR method: outliers as defined as values less than Q1 - 1.5*IQR or greater than Q3 + 1.5*IQR

# Compute the 1st and 3rd quartiles (Q1 and Q3), using the apply function by columns
Q1 <- apply(data_frame, 2, quantile, 0.25, na.rm=TRUE)
Q3 <- apply(data_frame, 2, quantile, 0.75, na.rm=TRUE)

# Compute the Interquartile range (IQR)
IQR <- Q3 - Q1

# Identify rows that contain at least one outlier, we use the apply function by rows to identify them
outlier_rows <- apply(data_frame, 1, function(x) any(x < (Q1 - 1.5*IQR) | x > (Q3 + 1.5*IQR)))

# Store outlier rows in a separate data frame
outliers_df <- data_frame[outlier_rows, ]

# Remove outliers by doing data_frame <- data_frame[! ... , ]
data_frame <- data_frame[!outlier_rows, ]

# 5. Summary statistics
summary(data_frame)

```
Now that we cleaned the data, we proceed to do the Exploratory Data Analysis. We used the steps provided in [this guide](https://www.kaggle.com/code/georgyzubkov/heart-disease-exploratory-data-analysis).  

The dataset contains 303 records, each variable has 15 fields. The dataset contains variables of different types:

**- Discrete:**  
  - `age`  
  - `trestbps` (resting blood pressure, in mmHg on admission to the hospital)  
  - `chol` (serum cholesterol levels)  
  - `thalach` (maximum heart rate achieved)  
  - `ca` (number of major vessels)  

**- Continuous:**  
  - `oldpeak` (ST depression induced by exercise relative to rest)  

**- Categorical:**  
  - `sex`  
  - `cp` (chest pain type)  
  - `fbs` (fasting blood sugar)  
  - `restecg` (resting electrocardiographic results)  
  - `exang` (exercise induced angina)  
  - `slope` (the slope of ST segment during exercise)  
  - `thal` (represents different types of thalassemia)  


```{r}
#XRS 1. Check the dataset for gaps in the data
# Create a data frame for the missing values for plotting
df_missing <- data.frame(Variable = names(missing_counts), Missing_Count = missing_counts)
# Load ggplot2
library(ggplot2)

# Plot
ggplot(df_missing, aes(x = reorder(Variable,-Missing_Count), y = Missing_Count)) +
  geom_bar(stat = "identity", fill = "darkblue", alpha = 0.7)+
  coord_flip() + # We flip the graph for improved readability
  labs(title = "Check for gaps in our data",
       x = "Variables",
       y = "Missing values counts") +
  theme_minimal()
```
We do have some missing values, but we removed them in the cleaning steps above.
```{r}
#XRS 2. Look at the distribution of the target variable: heart disease
num_counts <- as.data.frame(table(data_frame$num))

# Plot
ggplot(num_counts, mapping = aes(x=Var1, y=Freq)) +
  geom_bar(stat = "Identity", fill = "#6BACD0") +
  labs(title = "Distribution of our target variable",
       x = "num",
       y = "Count") +
  theme_minimal()
```
The sample seems to be unbalanced, with the majority of the patients being in the category 0 .
```{r}
# XRS 3: Indicate the types of variables (see paragraph above)
# XRS 4: Look at unique values in the categorical objects
unique(data_frame$sex)
unique(data_frame$cp)
unique(data_frame$fbs)
unique(data_frame$restecg)
unique(data_frame$exang)
unique(data_frame$slope)
unique(data_frame$thal)

# WB 5: Analyze numerical variables (visualizing continuous columns)
num_vars <- c("age", "trestbps", "chol", "thalach", "oldpeak")  
cat_vars <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "thal", "ca")  

for (var in num_vars) {
  print(
    ggplot(data_frame, aes_string(x = var)) +
      geom_histogram(fill = "blue", color = "black", alpha = 0.7, bins = 30) +  # Auto-adjust bins
      labs(title = paste("Distribution of", var), x = var, y = "Count") +
      theme_minimal()
  )
}
```
The distributions of age, cholesterol, and maximum heart rate appear roughly normal, while resting blood pressure shows multiple peaks, suggesting possibly different subgroups within the dataset. ST depression is highly right-skewed, with most values close to zero, indicating that significant ST depression is relatively rare in the dataset.
```{r}
# WB 6. Look at the distribution of number of people with heart disease from various factors
# Get the number of unique values in num
num_levels <- length(unique(data_frame$num))

# Generate a color palette with enough colors
color_palette <- scales::hue_pal()(num_levels)

for (var in num_vars) {
  print(
    ggplot(data_frame, aes_string(x = "factor(num)", y = var, fill = "factor(num)")) +
      geom_boxplot(alpha = 0.7, outlier.color = "red") +
      scale_fill_manual(values = color_palette) +  # Use dynamic colors
      labs(title = paste("Heart Disease vs", var), x = "Heart Disease (num)", y = var) +
      theme_minimal()
  )
}
```
Thalach decreases, while oldpeak increases with heart disease severity, suggesting their strong predictive value. Age shows a slight upward trend, while trestbps and chol display high variability but no clear pattern.
```{r}
for (var in cat_vars) {
  print(
    ggplot(data_frame, aes_string(x = var, fill = "factor(num)")) +
      geom_bar(position = position_dodge(width = 0.9)) +  # Improved positioning
      labs(title = paste("Heart Disease Distribution by", var), x = var, y = "Count") +
      theme_minimal()
  )
}
```
Heart disease is more prevalent in males, higher chest pain types, and abnormal thalassemia. Higher major vessels count (ca), slope, and exercise-induced angina (exang) also show trends linked to severity.
```{r}
# WB 7. Look at the distribution of categorical variables depending on gender
for (var in cat_vars) {
  print(
    ggplot(data_frame, aes_string(x = var, fill = "factor(sex)")) +
      geom_bar(position = position_dodge(width = 0.9)) +
      labs(title = paste(var, "Distribution by Gender"), x = var, y = "Count") +
      theme_minimal()
  )
}
```
Males dominate the dataset and have higher chest pain type, abnormal thalassemia, and major vessels (ca). Gender-based differences exist in exercise-induced angina and resting ECG, potentially impacting heart disease risk.
```{r}
# WB 8. Look at outliers in numerical values depending on several factors
for (var in num_vars) {
  print(
    ggplot(data_frame, aes_string(x = "factor(num)", y = var, fill = "factor(num)")) +
      geom_boxplot(alpha = 0.7, outlier.color = "red") +
      scale_fill_manual(values = color_palette) +
      labs(title = paste("Outliers in", var, "by Heart Disease"), x = "Heart Disease (num)", y = var) +
      theme_minimal()
  )
}
```
Outliers are present in cholesterol, maximum heart rate (thalach), and ST depression (oldpeak), particularly in severe heart disease cases. Oldpeak increases with disease severity, while thalach decreases, reinforcing their potential as key indicators.
```{r}
for (var in num_vars) {
  print(
    ggplot(data_frame, aes_string(x = "factor(sex)", y = var, fill = "factor(sex)")) +
      geom_boxplot(alpha = 0.7, outlier.color = "red") +
      scale_fill_manual(values = c("#00BFC4", "#F8766D")) +
      labs(title = paste("Outliers in", var, "by Gender"), x = "Gender (0 = Female, 1 = Male)", y = var) +
      theme_minimal()
  )
}

#CONCLUSION ---> e.g. from all the graphs presented, it can be concluded that e.g. "alcohol consumption and smoking are not the main factors in heart disease"


```
Outliers are minimal across genders, with males showing slightly higher oldpeak values and females exhibiting lower but more varied thalach. Other variables have similar distributions, suggesting gender differences may be less pronounced in outliers.


Heart disease severity correlates with increasing oldpeak and decreasing thalach, with males, higher cp, thal 7, and ca > 0 more affected. Chol and trestbps show variability but are weaker predictors, while outliers have minimal impact on trends. Lifestyle factors like alcohol and smoking were not assessed.

## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r}

library(corrplot)
library(dplyr)
# Visualize Correlation Matrix
# Select relevant columns
selected_data <- data_frame[, c("num", "oldpeak", "age", "trestbps", "chol", "thalach")]

# Ensure columns are numeric
selected_data <- selected_data %>%
  mutate(across(everything(), as.numeric))

#correlation matrix with the numeric measurements
cor_matrix <- cor(selected_data, method = "spearman")
corrplot(cor_matrix, method = "color", type = "full", addCoef.col = "black",
         tl.col = "black", tl.srt = 45, number.cex = 0.8)

```


# 2. Difference in mortality rates in hospitalized COVID-19 patients 
Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}
library(readxl)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)

table <- read_excel("data/Table 1.XLSX")

# Figure 1, part 1
# Standardize column names
table <- table %>%
  rename_with(~ gsub("\\s+", "_", .)) %>%
  rename_with(tolower)

# Step 1: Remove exact duplicate rows
table_unique <- table %>% distinct()    # Two rows were removed

# Further explore the data: check for duplicate ids in table_unique
duplicates_table_unique_1 <- table_unique %>%
  group_by(id) %>%
  filter(n() > 1)

```
We identified 6 IDs that appear twice in the dataset, with different values for their corresponding measurements.
Since we cannot determine which of the two entries is the correct or most reliable one, keeping them could introduce 
inconsistencies and potential bias in the analysis.
Rather than arbitrarily selecting one of the rows, we opt to remove these ambiguous cases to ensure data integrity.
```{r}
# Remove rows that share the exact same id
table_unique <- table_unique %>%
  group_by(id) %>%
  filter(n() == 1) %>%
  ungroup()

# Create an Age Group column for plotting
table_unique <- table_unique %>%
  mutate(age_group = cut(age, breaks = seq(20, 100, by = 10), right = FALSE, include.lowest = TRUE))

# Step 4: Create the histogram with adjustments for styling
ggplot(data = table_unique, mapping = aes(x = age)) +
  geom_histogram(
    binwidth = 10,
    color = "black",
    fill = "#E0EEEE",
    boundary = 20
  ) +
  scale_x_continuous(breaks = seq(20, 100, by = 10), limits = c(20, 100)) +
  labs(
    x = "Age (years)",
    y = "Frequency (n)",
    title = "Age"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

# Figure 1, part 2
# Define the table data
data <- matrix(c("G1", "-",  "-",   "-",
                 "G2", "-/+", "+/-", "-",
                 "G3", "+",   "-",   "+",
                 "G4", "-/+", "+",   "+"),
               nrow = 4, ncol = 4, byrow = TRUE)

colnames <- c("Clinical\nclassification", "NIV", "AMV", "ARDS")

# Create the tableGrob
table <- tableGrob(data,
                   rows = NULL, # No row names
                   cols = colnames)

# Define the theme with alternating row colors and centered text
theme <- ttheme_minimal(
  core = list(
    fg_params = list(hjust = 0.5, x = 0.5, fontsize = 12), # Center-align and font size
    bg_params = list(fill = c(rep(c("white", "lightgray"), length.out = nrow(data))), 
                     col = "white") # Alternating row colors with white gridlines
  ),
  colhead = list(
    fg_params = list(fontsize = 14, fontface = "bold"), # Header style
    bg_params = list(fill = "lightgray", col = "white") # Header background
  )
)

# Recreate the table with the theme
table <- tableGrob(data, rows = NULL, cols = colnames, theme = theme)

# Add white lines (gridlines) between rows and columns
table$grobs <- lapply(table$grobs, function(grob) {
  if (inherits(grob, "rect")) {
    grob$gp$col <- "white" # Set gridline color to white
    grob$gp$lwd <- 2       # Set line width
  }
  return(grob)
})

# Add a title
title <- textGrob("Definition of the clinical classification",
                  gp = gpar(fontsize = 14, fontface = "bold"))

# Combine title and table into a single plot with closer spacing
grid.newpage()
grid.arrange(table, ncol = 1)
```
```{r}
#Figures 1c and 1d
# Classify patients into G1, G2, G3, or G4
table_unique <- table_unique %>%
  mutate(
    group = case_when(
      use_of_niv == "No" & use_of_amv == "No" & days_with_ards == 0 ~ "G1",
      days_with_ards == 0 ~ "G2",
      use_of_niv == "Yes" & use_of_amv == "No" & days_with_ards > 0 ~ "G3",
      use_of_amv == "Yes" & days_with_ards > 0 ~ "G4"
    )
  ) %>%
  filter(!is.na(group))  # Remove NA values

# Count patients in each group
group_counts <- table_unique %>%
  count(group)

# Create Figure 1C - Clinical Classification Bar Plot
fig_1c <- ggplot(group_counts, aes(x = group, y = n, fill = group)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, size = 5, fontface = "plain") +  # Not bold
  scale_fill_manual(values = c("G1" = "#76c7c0", "G2" = "#f6e58d", "G3" = "#a29bfe", "G4" = "#ff7675")) +
  scale_y_continuous(breaks = seq(0, 80, by = 20), limits = c(0, 80)) +
  labs(title = "Clinical classification", x = "Clinical classification", y = "Frequency (n)") +  
  theme_classic() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 12, face = "plain"),  
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.line.x = element_blank(),  # Remove X-axis line
    axis.ticks.x = element_blank(),  # Remove X-axis ticks
    legend.position = "none",  # Remove legend
    plot.margin = margin(20, 40, 20, 40)  # Increase width spacing
  )

# Count number of deaths
vital_counts <- table_unique %>%
  count(death) %>%
  arrange(match(death, c("No", "Yes")))

# Create Figure 1D - Vital Status Bar Plot
fig_1d <- ggplot(vital_counts %>% filter(death %in% c("No", "Yes")), aes(x = death, y = n, fill = death)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = n), vjust = -0.2, size = 5, fontface = "plain") +  # Keep labels exactly on top
  scale_fill_manual(values = c("No" = "#76c7c0", "Yes" = "#f6e58d")) +
  scale_y_continuous(breaks = seq(0, 160, by = 50), limits = c(0, 160)) +  # Allow space for labels
  labs(title = "Vital status", x = "Death", y = "Frequency (n)") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 12, face = "plain"),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.line.x = element_blank(),  # Remove X-axis line
    axis.ticks.x = element_blank(),  # Remove X-axis ticks
    legend.position = "none",  # Remove legend
    plot.margin = margin(20, 40, 20, 40)  # Adjust spacing
  )


# Display 1c, and 1d
print(fig_1c)
print(fig_1d)

```


## Reproduce Figure 2 from the publication
but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(pheatmap)

# 1. Read the data from the Excel document
raw_data <- read_excel("data/Table 2.XLSX", 1)

# View the first few rows
head(raw_data)

# The first row contains a header
colnames(raw_data) <- raw_data[1, ] # Set the first row as column names
clean_data <- tail(raw_data,-1)      # Remove the first row

# For more clarity, make sure the first column is named "id"
colnames(clean_data)[1] <- "id"

# Check data types of the data frame
str(clean_data)           # All columns have type 'char'

# 2. Transform cytokine data to numeric type for further processing
numeric_columns <- colnames(clean_data)[2:ncol(clean_data)]                    # Select columns except for the first one (id)
clean_data[numeric_columns] <- lapply(clean_data[numeric_columns], as.numeric) # Convert to numeric type, use the lapply function to apply as.numeric to each element of the numeric_columns
# Verify the columns have the correct type
str(clean_data)

# 3. Handle multiple measurements per patient
# First of all, remove leading/trailing spaces in id
clean_data$id <- trimws(clean_data$id)

# Then, fill in the missing id values for rows with 'NA' using the fill() function, for the 
# data aggregation to work correctly
clean_data <- clean_data %>%
  fill(id, .direction = "down")

# Some patients have multiple measurements, taken on different days. To aggregate their cytokine levels, we choose to replace multiple measurements by their mean
aggregated_data <- clean_data %>%
  group_by(id) %>%
  summarise(across(everything(), ~mean(.x, na.rm = TRUE)))

# Merge patient group information (G1-G4)
aggregated_data <- left_join(aggregated_data, table_unique %>% select(id, group), by = "id")

# Scale each cytokine between 0 (min) and 1 (max)
#min_max_norm <- function(x) {
#  return((x - min(x, na.rm = TRUE))/(max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))
#}

# Apply normalization to all cytokine columns (except for id)
#normalized_data <- aggregated_data %>%
#  mutate(across(-c(id, group), min_max_norm))

# Remove unnecessary columns and prepare the data for heatmap
# Only keep the first 14 columns of the data frame, except column Day (which contain the cytokine measurements we are interested in plotting)
new_aggregated_data <- select(aggregated_data, c(id, colnames(aggregated_data)[3:14], group))

# Create annotation data frame for G1-G4 groups
annotation_col_group <- data.frame(Group = factor(new_aggregated_data$group))

# Define colors for the groups
group_colors <- list(Group = c("G1" = "green", "G2" = "blue", "G3" = "red", "G4" = "purple"))

# Make sure we are working with a data.frame and not a tibble
new_aggregated_data <- as.data.frame(new_aggregated_data)
# Set patient id as rownames
rownames(new_aggregated_data) <- new_aggregated_data$id

# Assign row names according to patient id
rownames(annotation_col_group) <- new_aggregated_data$id 

# Create heatmap
# Convert to numeric matrix for heatmap, preserve patient id as row names
data_matrix <- as.matrix(new_aggregated_data[,-c(1, ncol(new_aggregated_data))]) # Exclude patient id and group

# Generate the heatmap
pheatmap(t(data_matrix),
         color = colorRampPalette(c("#FDF5EB", "#FD9C50","#8D2A08"))(100),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         fontsize_row = 10,
         main = "Cytokine Expression Heatmap (Relative Expression)",
         annotation_col = annotation_col_group,  
         annotation_colors = group_colors)

```


```{r}

```


```{r}
```

## Improve figure 2 of the publication
Add a second annotation with information of deathm and a third one with information of gender

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(pheatmap)

# 1. Read the data from the Excel document
raw_data <- read_excel("data/Table 2.XLSX", 1)

# View the first few rows
head(raw_data)

# The first row contains a header
colnames(raw_data) <- raw_data[1, ] # Set the first row as column names
clean_data <- tail(raw_data,-1)      # Remove the first row

# For more clarity, make sure the first column is named "id"
colnames(clean_data)[1] <- "id"

# Check data types of the data frame
str(clean_data)           # All columns have type 'char'

# 2. Transform cytokine data to numeric type for further processing
numeric_columns <- colnames(clean_data)[2:ncol(clean_data)]                    # Select columns except for the first one (id)
clean_data[numeric_columns] <- lapply(clean_data[numeric_columns], as.numeric) # Convert to numeric type, use the lapply function to apply as.numeric to each element of the numeric_columns
# Verify the columns have the correct type
str(clean_data)

# 3. Handle multiple measurements per patient
# First of all, remove leading/trailing spaces in id
clean_data$id <- trimws(clean_data$id)

# Then, fill in the missing id values for rows with 'NA' using the fill() function, for the
# data aggregation to work correctly
clean_data <- clean_data %>%
  fill(id, .direction = "down")

# Some patients have multiple measurements, taken on different days. To aggregate their cytokine levels, we choose to replace multiple measurements by their mean
aggregated_data <- clean_data %>%
  group_by(id) %>%
  summarise(across(everything(), ~mean(.x, na.rm = TRUE)))

# Merge patient group information (G1-G4)
aggregated_data <- left_join(aggregated_data, table_unique %>% select(id, group), by = "id")

# Scale each cytokine between 0 (min) and 1 (max)
min_max_norm <- function(x) {
  return((x - min(x, na.rm = TRUE))/(max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))
}

# Apply normalization to all cytokine columns (except for id)
normalized_data <- aggregated_data %>%
  mutate(across(-c(id, group), min_max_norm))

# Remove unnecessary columns and prepare the data for heatmap
# Only keep the first 14 columns of the data frame, except column Day (which contain the cytokine measurements we are interested in plotting)
new_normalized_data <- select(normalized_data, c(id, colnames(normalized_data)[3:14], group))

# Create annotation data frame for G1-G4 groups
annotation_col_group <- data.frame(Group = factor(new_normalized_data$group))

# Define colors for the groups
group_colors <- list(Group = c("G1" = "green", "G2" = "blue", "G3" = "red", "G4" = "purple"))

# Make sure we are working with a data.frame and not a tibble
new_normalized_data <- as.data.frame(new_normalized_data)
# Set patient id as rownames
rownames(new_normalized_data) <- new_normalized_data$id

# Assign row names according to patient id
rownames(annotation_col_group) <- new_normalized_data$id 

# Create heatmap
# Convert to numeric matrix for heatmap, preserve patient id as row names
data_matrix <- as.matrix(new_normalized_data[,-c(1, ncol(new_normalized_data))]) # Exclude patient id and group

# Generate the heatmap
pheatmap(t(data_matrix),
         color = colorRampPalette(c("#FDF5EB", "#FD9C50","#8D2A08"))(100),
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         fontsize_row = 10,
         main = "Cytokine Expression Heatmap (Relative Expression)",
         annotation_col = annotation_col_group,  
         annotation_colors = group_colors)


```


# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
